---
title: "Regression Models Course Project"
author: "Katya Demidova"
date: "23 December 2015"
output: html_document
---

## Overview

The main goal of this project is to explore the relationship between a set of variables and miles per gallon (MPG) from the `mtcars` data set. We are particularly interested in the following two questions:

* "Is an automatic or manual transmission better for MPG"
* "Quantify the MPG difference between automatic and manual transmissions".

Short summary:

* Cars with manual transmission have significantlly more miles per gallon than cars with automatic transmission;
* Manual transmission is better for MPG by 1.8 compared to automatic transmission (adj. by `cyl`, `hp`, `wt`).

## Loading and exploring the data

```{r}
library(ggplot2)
library(datasets)
```

To answer these questions, we will use the `mtcars` data set. This data was extracted from the 1974 *Motor Trend* US magazine. It comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

The `mtcars` data set has 32 observations of 11 variables:
```{r}
str(mtcars)
```

In this work we will be interested in following variables: 
* `mpg`:	Miles/(US) gallon (numeric),
* `am`:	Transmission (0 = automatic, 1 = manual) (numeric).

```{r}
summary(mtcars$mpg)
```

The histogram of MPG can be seen in appendix.

```{r}
# factor the am variable for convenience
mtcars$am <- factor(mtcars$am, labels = c("automatic", "manual"))
# factor some other variables as well
mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
```

## Analysis

In our further analysis, we will focus on relationship between MPG and transmission (`mpg` and `am`), with `mpg` as the outcome and `am` as the predictor. As we can see in the boxplot of MPG by transmission type, there is an increase in MPG when transmission is manual comparing to automatic transmission:

```{r}
ggplot(aes(am, mpg, fill=am), data = mtcars) + 
  geom_boxplot() + 
  ggtitle("Boxplot of MPG by transmission type") + 
  ylab("MPG") + xlab("Transmission type") + 
  theme_update(legend.position=c(0.85, 0.15))
```

To find out if this increase is statistically significant, we will perform a t-test (assuming that the distribution is normal):
```{r}
t.test(mpg ~ am, data = mtcars)
```

With p-value of `0.001374`, we can see that this difference is significant.

### Linear regression model

```{r}
lin.model <- lm(mpg ~ am, data = mtcars)
summary(lin.model)
```

The intercept coefficient (`r coef(lin.model)[1]`) is the empirical mean of MPG for cars with automatic transmissions. The `am` coefficient (`r coef(lin.model)[2]`) is the mean increase of MPG for cars with manual transmissions. The mean MPG for cars with manual transmissions is `intercept` + `am` = `r sum(coef(lin.model))[1]+ coef(lin.model)[2]`.

The p-value is `0.000285`, which means that there is a significant difference in MPG between groups with different transmission. 

The adjusted R-squared is `r summary(lin.model)$adj.r.squared`, which means that only ~34% of the variability is explained by this model. This percentage is relatively small, so we should find out which of other variables could be predictors in our model.

### Multivariate regression model

To begin with, we will build a model with `mpg` as the outcome and all other variables as predictors. Next, we will use this model as the initial model in the stepwise search by `step` algorithm.

```{r results='hide'}
init.model <- lm(mpg ~ ., data = mtcars)
step.model <- step(init.model)
```
```{r}
summary(step.model)
```

The best model found by the `step` algorithm contains `cyl`, `hp`, `wt`, and `am` variables. 

```{r}
anova(lin.model, step.model)
```
We have compared this model with the linear model. From the results above, we can tell that there are significant differences between two models, in other words, `cyl`, `hp`, and `wt` variables contribute to the model.

So with these three variables as confounders, adjusted R-squared is `r summary(step.model)$adj.r.squared`, meaning 84% of the variability is explained.

* Is an automatic or manual transmission better for MPG?
* Quantify the MPG difference between automatic and manual transmissions.

## Appendix
```{r}
head(mtcars)
```
```{r results='hide'}
ggplot(aes(mpg), data = mtcars) + geom_histogram() + ggtitle("Histogram of MPG")
```