---
title: "Statistical Inference Course Project Part 2"
author: "Katya Demidova"
date: "10 December 2015"
output: html_document
---


In this part of the course project we're going to analyze the ToothGrowth data from the R `datasets` package. Description of this data set: the response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

```{r}
library(datasets)
tg <- ToothGrowth
str(tg)
```

The ToothGrowth data is a data frame with 60 observations of 3 variables:

* `len`: tooth length (numeric),
* `supp`: supplement type (factor with 2 levels `OJ` and `VC`, orange juice and ascorbic acid respectively),
* `dose`: dose of Vitamin C in miligrams (numeric).

```{r}
ggplot(data = tg, aes(as.factor(dose), len)) + geom_point(aes(colour=supp), size=4) + ggtitle("Tooth growth by dose of Vitamin C\n") + facet_grid(. ~ supp)
```

The plot above could help us with comparing tooth growth by `supp` and `dose`:
* it is clear from the plot that the higher the dose - the larger the teeth, in other words, it looks like there is a positive correlation between dose level and tooth length in guinea pigs;
* we might also assume that orange juice causes larger tooth growth than ascorbic acid, though it is not clear from the data

2. Provide a basic summary of the data.
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)
4. State your conclusions and the assumptions needed for your conclusions. 
Some criteria that you will be evaluated on
1. Did you  perform an exploratory data analysis of at least a single plot or table highlighting basic features of the data?
2. Did the student perform some relevant confidence intervals and/or tests?
Were the results of the tests and/or intervals interpreted in the context of the problem correctly? 
3. Did the student describe the assumptions needed for their conclusions?